---
title: "Airline Delays - 2015 Review"
author: "Gary W. Deinert"
date: "Monday, May 04, 2015"
output: html_document
---
```{r echo=FALSE, warning=FALSE}
library(ggplot2)
library(reshape2)
library(grid)
setwd("C:/Users/Gary/Desktop/Data Science Specialzation/Contract Projects/Airline delays")
filenm = "airline_delay_causes.csv"
flydata <- read.csv(filenm,header=TRUE)


```
## Synopsis.
#### In February of 2015, airline carriers at various airports were logged by their different types of delays: late arrivals, weather, security, cancellations, diverted flights, etc.  Our data shows the number of events as well as the total time lost due to the different types of delays.  

```{r echo=FALSE}

## step 1:  make the data tidy.  melt the df and recast matching one observation per row.
flights <- melt(flydata, id=c("year","month","carrier","carrier_name","airport","airport_name"))

## subset for differences between number of flights delayed and time of delays
counts <- flights[flights$variable %in% c("arr_flights","arr_del15","carrier_ct","weather_ct","nas_ct","security_ct","late_aircraft_ct","arr_cancelled","arr_diverted"),]
times <- flights[flights$variable %in% c("arr_delay","carrier_delay","weather_delay","nas_delay","security_delay","late_aircraft_delay"),]

## create summary of delay events by carrier and airport
airpt_counts <- aggregate.data.frame(counts$value,by=list(counts$airport,counts$variable),sum)
names(airpt_counts) <- c("airport","delay_type","number")
carrier_counts <- aggregate.data.frame(counts$value,by=list(counts$carrier_name,counts$variable),sum)
names(carrier_counts) <- c("airline","delay_type","number")

## create summary of delay times by carrier and airport
airpt_times <- aggregate.data.frame(times$value, by=list(times$airport,times$variable),sum)
names(airpt_times) <- c("airport","delay_type","time")
carrier_times <- aggregate.data.frame(times$value,by=list(times$carrier_name,times$variable),sum)
names(carrier_times) <- c("airline","delay_type","number")

## remove normal arriving flights from counts of delays
carrier_delays <- carrier_counts[carrier_counts$delay_type != "arr_flights",]
airpt_delays <- airpt_counts[airpt_counts$delay_type != "arr_flights",]

## too many airports for good reporting.  Limit to top 25 offenders.
hi_delay_airpts <- aggregate.data.frame(airpt_delays$number,by=list(airpt_delays$airport),sum)
names(hi_delay_airpts) <- c("airport","delay_events")
hi_delay_airpts <- hi_delay_airpts[order(hi_delay_airpts$delay_events,decreasing=TRUE),][1:25,]
```
### Exploratory Data Analysis.

```{r echo = FALSE, warning=FALSE, message=FALSE}
## Create the 'multiplot' function, allowing us to put multiple plots on one page using ggplot2
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

## plots by airline carrier ===================

p1 <- ggplot(data = carrier_delays, 
       aes(x=carrier_delays$airline,
           y=carrier_delays$number,
           fill=carrier_delays$delay_type,)) + 
      geom_bar(stat="identity", width=0.75) + 
      coord_flip() + 
      xlab("US Airline Carriers") +
      ylab("Number of Delays, February 2015") +
      scale_fill_brewer(palette="Accent") +
      scale_fill_discrete(name="Delay Type") +
      ggtitle("Delay Events by Airline\n") +
      theme(plot.title = element_text(lineheight = .8, face = "bold"),
            axis.text.x = element_text(angle=90))

p2 <- ggplot(data = carrier_times, 
       aes(x=carrier_times$airline,
           y=carrier_times$number,
           fill=carrier_times$delay_type,)) + 
      geom_bar(stat="identity",width=0.75) + 
      coord_flip() + 
      xlab("US Airline Carriers") +
      ylab("Time Delays (minutes), February 2015") +
      scale_fill_brewer(palette="Accent") +
      scale_fill_discrete(name="Cause of Delay") +
      ggtitle("Delay Amounts(Time) by Airline\n") +
      theme(plot.title = element_text(lineheight = .8, face = "bold"),
            axis.text.x = element_text(angle=90))
print(p1)
print(p2)

## plots by airport ========================================

sel_airpts <- airpt_delays[airpt_delays$airport %in% hi_delay_airpts$airport,]
p3 <- ggplot(data = sel_airpts, 
       aes(x=sel_airpts$airport,
           y=sel_airpts$number,
           fill=sel_airpts$delay_type,)) + 
      geom_bar(stat="identity", width=0.75) + 
      coord_flip() + 
      xlab("US Airports") +
      ylab("Number of Delays, February 2015") +
      scale_fill_brewer(palette="Accent") +
      scale_fill_discrete(name="Delay Type") +
      ggtitle("What causes delays at the top 25 delay airports?\n") +
      theme(plot.title = element_text(lineheight = .8, face = "bold"),
            axis.text.x = element_text(angle=90))

sel_airpts_time = airpt_times[airpt_times$airport %in% hi_delay_airpts$airport,]
p4 <- ggplot(data = sel_airpts_time, 
       aes(x=sel_airpts_time$airport,
           y=sel_airpts_time$time,
           fill=sel_airpts_time$delay_type,)) + 
      geom_bar(stat="identity",width=0.75) + 
      coord_flip() + 
      xlab("US Airports") +
      ylab("Time Delays (minutes), February 2015") +
      scale_fill_brewer(palette="Accent") +
      scale_fill_discrete(name="Cause of Delay") +
      ggtitle("How much time is lost by delay at the top 25 delay airports?\n") +
      theme(plot.title = element_text(lineheight = .8, face = "bold"),
            axis.text.x = element_text(angle=90))
print(p3)
print(p4)
